---
title: "PCA Analysis Cropland Cover"
author: "Oswaldo Felix"
date: "2020-02-05"
output: html_document
tags:
  - statistical
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = FALSE, warning = FALSE)
```

## Introduction

Principal-component-analysis is extremely useful for discerning correlation of variables within a given dataset. This project seeks to find country-level effects of increasing cropland on annual environmental extremes. By plotting eigenvectors of each variable on a biplot, one can quickly qualitatively assess trends among variables.

```{r}
# Load libraries

library(tidyverse)
library(here)
library(janitor)
library(ggfortify)

# Read data

world_df <- read_csv(here::here("csv", "world_env_vars.csv")) %>% 
  clean_names() %>%
  rename("rain_seasonality" = "rain_seasonailty") %>% 
  select(-country:-slope, -isothermality, -wind, -cloudiness,
         -contains("range"), - contains("seasonality"))

```

```{r}
# Create subset

world_subset <- world_df %>% 
  drop_na()

my_world_pca <- prcomp(world_subset, scale = TRUE)

```


### PCA Biplot
```{r,}
# Generate biplot

my_biplot <- autoplot(my_world_pca,
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE) +
  scale_x_continuous(limits = c(-0.15, 0.15)) +
  theme_light()

my_biplot
```

**Figure 1.** *All variables show some negative correlation (angle > 90) with cropland cover*

Takeaways:

- Cropland cover is strongly inversely correlated with cold temperature statistics

- Cropland cover is also inversely correlated with large amounts of rain, but not as strongly as the relationship seen with cold temperatures

- Tree canopy cover positively correlates with rain statistics

- Rain and temperature statistics correlate within respective categories

### Citations

The data was compiled by @zander_venter on Kaggle and originates from the [Google Earth Engine](https://earthengine.google.com/).